---
title: "02_DataAnalysis"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

# Data Analysis 

Run setup if starting from a clean environment.

```{r}

source("01_Setup.R")

#Read in the clean datasets
in.BCCWS<-read.csv("Data/in.BCCWS.csv")
event.BCCWS<-read.csv("Data/event.BCCWS.csv")

#Manually Specify the start and end year of the analysis
Y1 = 1999
Y2 = 2023

```


## 2 Methods

Our analysis will use spatially varying coefficient models (SVCs, Gelfand et al. 2003) to account for relationships between variables that are not uniform across large spatial areas. This modelling approach was first applied to continent wide bird abundance data to assess winter bird population trends using discrete aerial units (Meehan et al. 2019) and an intrinsic conditional autoregressive model (iCAR; Besag 1974). The modelling framework was later adapted (Meehan et al. 2024) to incorporate continuous space using a triangulated model mesh and stochastic partial differential equation (SPDE; Lindgren et al. 2022). The SPDE models benefits include (1) finer resolution of trends, (2) a better understanding of the range of spatial correlation, and (3) a reduction in boundary effects associated with discrete-space analyses (e.g., iCAR). 

The basic statistical unit for the analysis is the count of each species on a survey route during each month of the BCCWS (September to April). Parameters include a site-specific random intercept (alpha) and site-specific random slope (tau) coefficients both with spatial SPDE structure, which allows for information on relative abundance and year effects, respectively, to be shared across neighbours. Year is scaled such that the maximum year is 0. Differences in relative abundance among routes (κappa) is modeled with an independent and identically distributed (idd) random effect. We include smooth effect (ar1) for day-of-year (doy_idx) to account for the seasonal distribution of counts, and DurationInHours as an covariate to account for differences in survey effort. The basic model formula is as follows:

Observation Count ~ DurationInHours + f(kappa, model="iid", hyper=hyper.iid) + f(doy_idx, model = "ar1", hyper=prec.prior) + f(alpha, model =spde)+ f(tau, model =spde)

Models are fit using a Bayesian framework with Integrated Nested Laplace Approximation (INLA) using the R-INLA package (Rue and Martino 2009) for R statistical computing (version 4.3.1; R Core Team 2021). Following model analysis and validation, posterior medians and 95% credible intervals (CI) are computed.

Trends: Multiple metrics can be used to estimate a 'trend' from the estimated population trajectories. We provide both 'slope' and 'endpoint' trends calculated post-hoc from the population trajectories. Linear slope trends provide an estimate of the medium and longer-term patterns of population change, removing the effect of the random annual fluctuations. Endpoint trends follow methods outlined in the bbsBayes package (Edward and Smith 2021). Specifically, endpoint trend estimates are the geometric mean annual change (% per year) in the population trajectory, using an interval-specific ratio of the smoothed annual indices at the start and end of the time period (Sauer and Link 2011). 

Trend estimates and 95% credible intervals were back-transformed to annual rates of population change using 100\*exp(estimate)-1. Trend were calculated using the full dataset, as well as for the most recent 10-years, and three generations (Bird et al. 2020), the latter of which are informative to status assessments in Canada (e.g., see COSEWIC's Quantitative criteria and guideline for the status assessment of wildlife species). If the three-generation length was less than 10 years, the output reflects the 10-year trend.

All model outputs are displayed on the NaureCounts web portal and are accessible through the naturecounts R package. The outputs from this analysis will therefore provide resource managers with openly accessible annual indices of abundance for model-based management planning.


### 2.1 Species Analysis
You can manually create the species list or use the unique species in the data set.

The code automatically creates output tables in the formate required for the State of Canada's Birds upload into NatureCounts. 

You can select the minimum data requirements for the analysis. The ones selected here were done following the inspection of the full data set and likely work well for most species. However, finer scale assessment may need to assess their suitability. 

```{r}

#Create a list of species
sp.list<-unique(in.BCCWS$SpeciesCode) 

#to restart the list if it crashes 
#sp.list<-sp.list[47:length(sp.list)]

#Load your saved events data which is needed for zero-filling
event.BCCWS<-read.csv("Data/event.BCCWS.csv")

#Define your site. 
site <- "BCCWS"

#Create a data frame to store the results. 
#This is the template used for the State of Canada's Birds and is required for upload into NatureCounts
source("03_OutputTables.R")

#Across routes with at least one detection, each species needs an overall abundance per year > 10, need to be detected in > one-half of the survey years, and on > 10 survey routes). Users may choose to adjust these parameters. 

min.abundance <- 50
min.years <- (Y2-Y1)/2
nroutes <- 10

#Now we can initiate the analysis.  
source("04_Analysis.R")

```

### 2.2 Guild Analysis

To be complete at a later date. 


